---
title: "AIM RSF R series: Data wrangling with `dplyr` and `tidyr`"
subtitle: "Based on Data Carpentry: R for Social Scientists"  
author: 
  - "Eirini Zormpa"
date: "15 November 2022 (last updated `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)


library(knitr)

library(RefManageR)
library(bibtex)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
bib <- ReadBib("bibliography.bib", check = FALSE)

library(readr)
library(here)
library(magrittr)
library(dplyr)
library(tidyr)
library(lubridate)

covid_data <- read_csv(here("data", "covid_data.csv"))
covid_dates <- read_tsv(here("data", "dates.tsv"))

covid_data_dates <- full_join(covid_data, covid_dates, by = "year_week")
                        
```

# Summary of session 2: Starting with data in R

- `r emo::ji("white_check_mark")` Understand what tidy data are and why it's a useful format.
- `r emo::ji("white_check_mark")` Read data into R.
- `r emo::ji("white_check_mark")` Understand and manipulate `data frames`.
- `r emo::ji("white_check_mark")` Understand and manipulate `factors`.

---

# Learning objectives: Data wrangling with `dplyr` and `tidyr`

- `r emo::ji("white_check_mark")` Subset columns or rows with `select` or `filter` and create new columns with `mutate`.
- `r emo::ji("white_check_mark")` Link the output of one function to the input of another function with the ‘pipe’ operator `%>%`.
- `r emo::ji("white_check_mark")` Combine datasets using `join`.
- `r emo::ji("white_check_mark")` Reshape a dataframe from long to wide format with the `pivot_wider` function.
- `r emo::ji("white_check_mark")` Export a dataframe to a .csv and .tsv file.

---
class: center, middle, inverse

# Exercise 1

`r emo::ji("clock")` **5 mins**

Subset the `covid_data` tibble such that you keep only observations from the TESSy COVID-19 `source` and retain only the variables `country`, `indicator`, `year_week` and `weekly_count`.


`r countdown::countdown(minutes = 5, color_running_background  = "lightgreen", color_running_text = "black", color_finished_background = "red", color_finished_text =  "grey30")`

---
class: center, middle, inverse

# Exercise 1 solution

```{r, eval=FALSE}
covid_data %>% 
  filter(source == "TESSy COVID-19") %>% 
  select(country, indicator, year_week, weekly_count)
```

--

Note that if you `select` before you `filter`, your code won't run.
That's because you're not retaining the variable that you use in your filtering.
When piping, **order matters**!

---
class: center, middle, inverse

# Exercise 2

`r emo::ji("clock")` **10 mins**

Create a new tibble `deaths_2021` that contains the total deaths for each country in 2021, arranged such that the country with the most deaths is at the top.


`r countdown::countdown(minutes = 10, color_running_background  = "lightgreen", color_running_text = "black", color_finished_background = "red", color_finished_text =  "grey30")`

---
class: center, middle, inverse

# Exercise 2 solution

```{r}
deaths_2021 <- covid_data_dates %>% 
  drop_na(weekly_count) %>% 
  mutate(year = year(from_date)) %>% 
  filter(year == 2021,
         indicator == "deaths") %>% 
  group_by(country) %>% 
  summarise(yearly_deaths = sum(weekly_count)) %>% 
  arrange(desc(yearly_deaths))
```

---

# Summary of packages we used today

- `r emo::ji("white_check_mark")` `readr` to read data into R and export it
- `r emo::ji("white_check_mark")` `dplyr` for a grammar of data manipulation
- `r emo::ji("white_check_mark")` `tidyr` to get data where variables are columns, observations are rows, and cells contain single values
- `r emo::ji("white_check_mark")` `magrittr` to get access to the `tidyverse` pipe `%>%`
- `r emo::ji("white_check_mark")` `stringr` to manipulate strings/characters
- `r emo::ji("white_check_mark")` `lubridate` to manipulate dates

---
class: center, middle

# Thank you for your attention `r emo::ji("sparkles")` `r emo::ji("pray")`

## See you next week for data visualisation with `ggplot2` `r emo::ji("artist_palette")`
